---
title: "2019.1.28 - Class 1 Data Viz"
author: "Michael Weisner"
date: "1/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started with GGPlot2

### Data Growth

If we take all the data between the beginning of time and 2008, we'll soon create that amount every minute.

So it's important we capture it!

## Methods of Capturing Data Trends:

### Modeling
Modeling is one way to understand relationships in data.

### Graphing / Visualization
This let's us quickly understand data in a more easily digestible way
Can also be good for digesting live-streaming data!

## Visualization
"Action or fact of visualization; the power or process of forming a mental picture or vision of something not present to the sight."

## Types of Visualization

1. To Understand/Discover Patterns (find things that rows don't reveal)
2. To Analyze (like seeing train slowdowns)
3. To Present (to make our point)

## Course Outline

1. Intro and Plotting with ggplot2
2. Working with Spatial Data. Visualizing Maps
3. Text Analysis and Visualization
4. Networks
5. Dynamic and Interactive Visuals

### ggplot2

+ Week 1
+ + Intro
+ + Basics 
+ + Concepts, data, aesthetics, geometries
+ Week 2
+ + Stats, coordinates, facets, themes
+ Best Practices
+ Week 3
+ + Refining Plots, themes, publication-ready
+ + Advanced plots and ggplot2 internals
+ Week 4
+ + Making ggplot interactive
+ + Using plot.ly

### Spatial Data

Maps are powerful!

### Text as Data

Text into data takes some steps... so we should go through those!

### Networks

e.g. social networks.

### Interactivity

Shiny! And "Data Driven Documents" (or D3) for javascript on the web.

## Data Visualization as a Skill
Data Visualization is typically a weak skill for data driven professionals (let's change it)

## RMarkdown
Use it for notes and homework!

It's clear and reproducible.

## Assignments

1. Final Report (30%) - a final group project in the form of a website and class presentation
2. Assignments (60%) - 4 individual assignments
3. Class Participation (off and online) & commentary on other student project (10%)

Final Project idea: Realtime MTA slowdowns with mapping of tweets with sentiment analysis?

## What makes for GOOD Data Visualization

Remember, there are two types of data visualization

1. Exploring (for data and researchers)
2. Explaining (for readers)


### Chart Examples

+ Bar Graphs tell us some relative differences
+ Pie Charts don't tell us the total, just relative comparisons.

### Bad Visualization:

1. Misleading
2. Wrong Form
3. Too Much or Too Little Information
4. Poor Design

#### Circles

Circles stink for comparisons. We can't easily compare areas between circles.

### Good Visualization:

1. Truthful (reliable and honest information)
2. Functional (organized that enables exploration)
3. Beautiful (at least attractive/intriguing/pleasing)
4. Insightful (reveals evidence that we have a hard time seeing otherwise)

### Other Tips

+ 3d and circular displays are rarely smart
+ For now let's practice making graphs!

## Starting with ggplot2!

### Packages
```{r}
library(ggplot2)
library(gapminder)
gapminder
```

Don't forget the `str()` command to look at structure of data.
```{r}
str(gapminder)
```

### Scatterplot
```{r}
gg_scatter <- ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) # nothing to plot yet
```

Above is just our plane, there's no data to be plotted yet. The reason is ggplot2 wants to make everything done in the same syntactical way.

Also note that we can assign ggplots to objects! This let's us modularly build plots.

```{r}
gg_scatter + geom_point()
```

#### Log Transformation

It looks like a logarythmic relationship so let's transform it
```{r}
p <- ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp))
p + geom_point() + scale_x_log10()
```

### Color
```{r}
p + 
  geom_point(aes(color = continent), alpha = 0.70) +
  scale_x_log10()
p
```

### Fitted Curves
```{r}
p + 
  geom_point(aes(color = continent), alpha = 0.70) +
  scale_x_log10() + 
  geom_smooth()
```

We can change color and take out standard error from the default loess curve.
```{r}
p + 
  geom_point(aes(color = continent), alpha = 0.70) +
  scale_x_log10() + 
  geom_smooth(color = "black", lwd=2, se = FALSE)
```

Here's a linear model
```{r}
p + 
  geom_point(aes(color = continent), alpha = 0.70) +
  scale_x_log10() + 
  geom_smooth(method="lm")
```

### Smooth Fit by Continent
```{r}
p + 
  geom_point(aes(color = continent), alpha = 0.70) +
  scale_x_log10() + 
  geom_smooth(lwd = 2, se = FALSE, aes(color = continent))
```

### Facet Wrapping

```{r}
p + 
  geom_point(aes(color = continent), alpha = 0.70) +
  scale_x_log10() + 
  geom_smooth(lwd = 2, se = FALSE, aes(color = continent)) + facet_wrap(~ continent)
```

## Grammar of Graphics

ggplot is based on a "grammar" of graphics, originated with Wilkinson (2005)

### Main Principles:

1. Graphics = distinct layers of grammatical elements
2. If necessary some additional rules about scales, projections in a coordinate system, and data transformations are possible
3. Plots arise through aesthetic mapping
4. The grammar produces "sentences" (mappings of data to objects) but they can be easily garbled if you define poor mappings

### Three Key Elements
 
1. Data (gapminder)
2. Aesthetics (gdp and life expectancy)
3. Geometries (points)

Also:

1. Data
2. Aesthetics
3. Geometries
4. Facets (facet by continent)
5. Statistics (add geom smooth or lm)
6. Coordinates (map?)
7. Themes (do some cleanup)





