---
title: "2019.2.11 - Class 3"
author: "Michael Weisner"
date: "2/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Agenda

1. Gestalt Principles - grouping items in a vizualization
2. ggplot2
+ additional geoms()
+ changing the details of plots
3. Some design principles from Edward Tufte

## Assignment 1
Febrary 25th at 5 pm!

Remember, check online resources first. Come to TAs when stumped.

## Final Project
3-4 People create an interactive website with accompanying project book.

1. Static images
2. Maps
3. Visualizations of text analyses
4. Network Visualizations

Prepare a hosted, interactive display of some visualizations.

### Data

1. be larger dataset to allow analysis in different forms
2. contain at best one or more of the following:
+ some geographic identifiers (cities, addresses, lat-long) to allow mapping
+ some text to allow simple text analyses
+ contain a network matrix to plot visual networks

Preference is for data that you either put together yourself from existing datasets or using some API to collect data you need.

### Output
Your final output should be something that you could use as a porfolio for job applications to show skills

### Hypothetical Example

+ __Title__: Unhealthy habits - bad diets are the root of obesity epidemic
+ __Abstract__: Obesity epidemic in US... bad diets are to blame... explore how have diets changed over time... how are diets related to socio-economic factors
+ __Possible Data__:
+ + adult and changed obesity rates in the U.S. (CDC) and the World (WHO) pair with geographic infor
+ + Pew Research surveys on food and nutrition attitudes and how diets have changed
+ + Content of Recipes changed over time (text analysis of recipes for measurements)
+ + Spread of Obesity in a large social network
+ __Process Book__: create an accumulation of how you got to your end goal ( could even include pictures of hand drawn ideas).

## Slides
+ ioslides are html5 slides.
+ Xaringan based on remark.js is good for picture rich slides.
+ reveal.js package offers good looking slides and nice transitions
+ Beamer is like latex.

## Gestalt Principles


### Simplicity 
Human brain loves simplicity, and tends to process simple patterns that are regular, even, and orderly faster than patterns that are more complex.

### Similarity
We identify things that are similar.

### Proximity
Elements close to each other are perceived as part of one group. We group things.

### Connectedness
We like things that are connected as grouped

### Continuity
Smooth lines are easier to follow

### Completion
We automatically and spontaneously perceive a full circle behind a square.

### Figure Ground
We can easily think of figures as what's in the foreground rather than the background. When those aren't obvious is when we have trouble.

### Common Fate
Elements with the same moving direction are perceived as a unity

### Takeaways

- **Similarity**: Keep it simple. Arrange data logically wherever possible. Use similar characteristics (color, size, shape, etc.) to establish relationships and to encourage groupings of objects.
- **Proximity**: Know what your chart’s information priority is, and then create groupings through proximity to support that priority.
- **Connection**: Use connected lines, shapes, and color regions as visual guides.
- **Continuity**: Arrange objects in a line to facilitate grouping and comparison.
- **Completion**: Use our tendency to visually complete shapes to reduce clutter. 
- **Figure/Ground**: Ensure there is enough contrast between your foreground and background so that charts and graphs are more legible.
- **Common Fate**: Use direction and/or movement to establish or negate relationships.

## Edward Tufte

Napoleon's Failure example.

"Clutter and confusion are not attributes of information [but rather] failure of design" - Edward Tufte

So, let's remove non-data "clutter" and elevate the data from the noise.

But it can't be ALL the clutter, because visual pieces can be good to make it more eye catching, draw your attention to important parts, and ultimately make it memorable.

### Tufte's Integrity and Design Principles

#### Strive for **graphical integrity**. Visual representations of data must tell the **truth**.

Graphical effects should be proportional to the numerical quantities

#### Clear, detailed, and thorough labeling and appropriate scales

#### Show Data variation, not Design variation.

#### in time-series displays of money, defalated and standardized units of monetary measurements are nearly always better than nominal units.

#### Maximize data=ink ratio

#### Avoid Chart Junk

#### Increase Data Density

Proportion of the total size of the graph that is dedicated to displaying data

Data density of a graphic = (number of entires in a data matrix) / (area of data graphic)


Refer to Tufte's graph about what kind of graph to use (relationship, comparison, distribution, composition).


## Applying Tufte to ggplot2

Sometimes you can't avoid tricks or hacks.

### Example Line Plot

```{r}
library(ggplot2)
x <- 1967:1977
y <- c(0.5,1.8,4.6,5.3,5.3,5.7,5.4,5,5.5,6,5)
d <- data.frame(x, y)
ggplot(d, aes(x,y)) + geom_line()
```

### Minimal Line Plot
Remember, you can make your own themes!
```{r}
ggplot(d, aes(x,y)) + geom_line() + geom_point(size=3) + 
  #theme_tufte(base_size = 15) + 
  theme(axis.title=element_blank()) +
  geom_hline(yintercept = c(5,6), lty=2) + 
  scale_y_continuous(breaks=seq(1, 6, 1), 
                     label=sprintf("$%s",seq(300,400,20))) + 
  scale_x_continuous(breaks=x,label=x) +
  annotate("text", x = c(1977,1977.2), y = c(1.5,5.5), 
           adj=1, family="serif", label = c(
          "Per capita\nbudget expenditures\nin constant dollars", 
          "5%"))
```

### Bar Plot

```{r}
library(psych)
library(tidyverse)
dat <- colMeans(msq[,c(2,7,34,36,42,43,46,55,68)], na.rm = T)*10

d <- tibble(trait = names(dat), value = dat)

ggplot(d, aes(x=trait, y=value)) + 
  geom_bar(stat = "identity")

```


### **Minimal** Bar Plot

```{r}
ggplot(d, aes(x=trait, y=value)) + 
  #theme_tufte(base_size=14, ticks=F) + 
  theme(axis.title=element_blank()) +
  geom_bar(width=0.25, fill="gray", stat = "identity") + 
  scale_y_continuous(breaks=seq(1, 5, 1)) + 
  geom_hline(yintercept=seq(1, 5, 1), col="white", lwd=1) +
  annotate("text", x = 3.5, y = 5, adj=1,  family="serif",
  label = c("Average scores\non negative emotion traits
          from 3896 participants\n(Watson et al., 1988)"))
```


### Boxplot


```{r}
ggplot(quakes, aes(factor(mag),stations)) +
  geom_boxplot()
```

### **Minimal** Boxplot


```{r}
ggplot(quakes, aes(factor(mag),stations)) + 
  #theme_tufte() +
 # geom_tufteboxplot(outlier.colour="transparent") + 
  geom_boxplot() +
  theme(axis.title=element_blank()) +
  annotate("text", x = 8, y = 120, adj=1,  family="serif",
  label = c("Number of stations \nreporting 
            Richter Magnitude\nof Fiji earthquakes (n=1000)"))
```

### Slopegraph
```{r error=TRUE}
library(RCurl)
# Get some functions for the slope graph
source_url("https://raw.githubusercontent.com/jkeirstead/r-slopegraph/master/slopegraph.r")
# The data is here
d <- read.csv(text = getURL("https://raw.githubusercontent.com/jkeirstead/r-slopegraph/master/cancer_survival_rates.csv"))
head(d)
```

```{r error=TRUE}
# Using James Keirstead’s GitHubs functions
# https://github.com/jkeirstead/r-slopegraph
## Convert raw data to right format

df <- build_slopegraph(d, x="year", y="value", 
                       group="group", method="tufte", 
                       min.space=0.04)
df$x <- df$variable
df$y <- df$value
df <- transform(df, x=factor(x, levels=c(5,10,15,20), 
         labels=c("5 years","10 years","15 years","20 years")), 
         y=round(y))
```

```{r error=TRUE}
plot_slopegraph(df) + 
  labs(title = "Estimates 
       of % survival rates") +
  theme_tufte(base_size=16, 
              ticks=F) + 
  theme(axis.title = 
          element_blank())

```

### Sparklines
```{r error=TRUE}
# Read in Data
d <- read_csv("data/us_nr_of_crimes_1960_2014.csv")
d <- tidyr::gather(d, Crime.Type = "Total":"Vehicle-Theft")
names(d) <- c("Year","Crime.Type","Crime.Rate")

# Get summary statistics we want to highlight in Sparkline Graph
d$Crime.Rate <- round(d$Crime.Rate,0)
mins <- group_by(d, Crime.Type) %&gt;% slice(which.min(Crime.Rate))
maxs <- group_by(d, Crime.Type) %&gt;% slice(which.max(Crime.Rate))
ends <- group_by(d, Crime.Type) %&gt;% filter(Year == max(Year))
```

```{r error=TRUE}
## Get quartile ranges for plot
d <- d %&gt;% group_by(Crime.Type) %&gt;%
  mutate(quart1 = quantile(Crime.Rate, probs = 0.25),
         quart2 = quantile(Crime.Rate, probs = 0.75))
d
```

```{r error=TRUE}
# Lot's of options - that is the "hacky" part
ggplot(d, aes(x=Year, y=Crime.Rate)) + 
  facet_grid(Crime.Type ~ ., scales = "free_y") + 
  geom_ribbon(aes(ymin = quart1, max = quart2), fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Crime.Rate), vjust = -1) +
  geom_text(data = maxs, aes(label = Crime.Rate), vjust = 2.5) +
  geom_text(data = ends, aes(label = Crime.Rate), 
            hjust = 0, nudge_x = 1) +
  geom_text(data = ends, aes(label = Crime.Type), 
            hjust = 0, nudge_x = 5) +
  expand_limits(x = max(d$Year) + 
                  (0.25 * (max(d$Year) - min(d$Year)))) +
  scale_x_continuous(breaks = seq(1960, 2010, 10)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 15, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())

```

## Economist Graph in ggplot2
Let's try to re-create [this graph](http://www.economist.com/node/21541178) using ggplot2:


