---
title: "2019.2.4 - Class 2"
author: "Michael Weisner"
date: "2/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ggplot2 is even used by BBC!

Visualization of the color of native american mourning --> picture of colors associated with various feelings in different cultures.

Not super great in terms of function and thus insight, but was truthful and relatively beautiful.

## Contrary Views

Monthly spirals by years?

Functional: We want to have legible functional graphs that provide the correct information easily to the interested reader and allow exploration.
Insightful: We want to reveal evidence that was otherwise hard to see.

SO

We must understand the basic principles of visual perception...

## Georgia Lupi
Check her out
https://medium.com/accurat-studio/beautiful-reasons-c1c6926ab7d7

## Perception

* We more easily see edges, contrasts, and movement
* We judge relative difference rather than absolute
* We process visual info using preattentive attributes and patterns
* We use gestalt principles to group individual elements into groups and differentiate patterns
* The elementary perceptual tasks involved determine the information we can easily estimate from graphs

## Preattentive Visual Properties

There are some formats we have a very easy time reading.

### Properties:

* A preattentive visual property is processed in spatial memory without our conscious action (but we don't need to be attentive)
* fast: < 500 milliseconds for the eye and brain to process
* color and contrast between features matters most

### Using Them
* Not all equally strong... 
* Do not use multiple preattentive channels for competing information
* For highlighting, use whatever feature dimension is used least in other parts of the design

### Making Things Stand Out

* Use these insights to design plots to emphasize particular pieces of information

**TAKEAWAYS:**

* Use color and shape sparingly to make the important info pop.
* Maximize the contrast with the background if the outlines of shapes are important.

### Symbols

Make symbols as distinct from each other as possible (and from their backgrounds). Redundant properties (color and shape) can be helpful.

### Elementary Perceptual Tasks

Some visual estimations are more easily carried out than others (remember pie charts are tough to read)

* Common scales are better than non-aligned scales.
* Length judgements are easier with a common baseline.
* Angles are often misjudged. Angles >90° are often overestimated, angles <90° often underestimated.
* Area judgments are less accurate than length judgments or position judgments.
* Volumes are misjudged even more. This is one reason 3-D charts are usually terrible.
* Slopes and changes in slope are prone to misjudgement.
* Colors are best reserved for displaying categories.


## The Plot-Making Process in GGPlot

| Element | Description | 
|------:|:-----:|
| **Data** | The dataset being plotted. |
| **Aes**thetics | The scales onto which we map our data. |
| **Geom**etries | The visual elements used for our data. | 

```{r}
library(ggplot2)
library(gapminder)
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) + 
        scale_x_log10() +
        geom_point(alpha=0.5, size=3, color="red")
```

```{r}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color=continent)) +
        geom_point() + scale_x_log10()
```

### Typical Aesthetics

| Aesthetic | Description | 
|------:|:-----:|
| x | X axis position |
| y | Y axis position |
| colour | Colour of dots, outlines of other shapes |
| fill | Fill colour |
| size | Diameter of points, thickness of lines |
| alpha | Transparency |
| linetype | Line dash pattern |
| labels | Text on a plot or axes |
| shape | Shape |

### Aesthetics & Geom

- each `geom()` layer allows you to set the aesthetics that make sense for the particular plot `geom()`
- for example, `geom_point` understands the following aesthetics: x, y, alpha, color, fill, group, shape, size, stroke. For `geom_point()` the aesthetics x and y are required.
- some aesthetics are limited to continous variables, others to categorical variables

### Aesthetics - Continuous Variables

Makes sense for continuous aesthetics, like color, size, alpha.

## Aesthetics - Categorical Variables - Mapping onto shape

```{r, fig.width=6, fig.height=4}
ggplot(d, aes(x = gdpPercap, y = lifeExp, shape=continent)) +
  scale_x_log10() + geom_point(alpha=0.3, size=4)
```

## Adding redundant channel to emphasize

```{r, fig.width=6, fig.height=4}
ggplot(d, aes(x = gdpPercap, y = lifeExp, shape=continent)) +
  scale_x_log10() + geom_point(alpha=0.3, size=4) +
  geom_point(data=filter(d, continent=="Americas"), 
   color="red", alpha=0.5, size=4) + theme(legend.position="none")
```

## Encircle to emphasize 
```{r, echo=FALSE}
previousplot <- ggplot(d, aes(x = gdpPercap, y = lifeExp, shape=continent)) +
  scale_x_log10() + geom_point(alpha=0.3, size=2) +
  geom_point(data=filter(d, continent=="Americas"), 
   color="red", alpha=0.5, size=2) + theme(legend.position="none")
previousplot
```


```{r, fig.width=6, fig.height=4}
# ggalt() includes the encircle() function
# devtools::install_github("hrbrmstr/ggalt", force=FALSE)
library(ggalt)
previousplot + geom_encircle(data=filter(d, country=="United States"), 
                expand=0.05, color="blue", linetype=2, size=2)
```

## Connect to emphasize 

```{r, echo=TRUE, fig.width=12, fig.height=8, eval=FALSE}
library(ggthemes)
ggplot(d, aes(x = gdpPercap, y = lifeExp, shape=continent)) +
  scale_x_log10() + 
  geom_path(data=filter(d, country=="United States"), 
            color="light blue", linetype=1, size=6) +
  geom_path(data=filter(d, country=="Venezuela"), 
            color="light green", linetype=1, size=6) +
  geom_path(data=filter(d, country=="Haiti"), 
            color="orange", linetype=1, size=6) +
  geom_point(alpha=0.3, size=4) +
  geom_point(data=filter(d, continent=="Americas"), 
              color="red", alpha=0.5, size=4) + 
  theme(legend.position="none") +
  annotate("text", x = c(40000), y = c(73), family="serif", size=6, 
           color="dark blue", label = c("United States")) +
  annotate("text", x = c(13000), y = c(63), family="serif", size=6, 
           color="dark green", label = c("Venezuela")) +
  annotate("text", x = c(1200), y = c(62), family="serif", size=6, 
           color="dark orange", label = c("Haiti")) + theme_tufte()
```

## Connect to emphasize 

```{r, echo=FALSE, fig.width=12, fig.height=8}
library(ggthemes)
ggplot(d, aes(x = gdpPercap, y = lifeExp, shape=continent)) +
  scale_x_log10() + 
  geom_path(data=filter(d, country=="United States"), 
            color="light blue", linetype=1, size=6) +
  geom_path(data=filter(d, country=="Venezuela"), 
            color="light green", linetype=1, size=6) +
  geom_path(data=filter(d, country=="Haiti"), 
            color="orange", linetype=1, size=6) +
  geom_point(alpha=0.3, size=4) +
  geom_point(data=filter(d, continent=="Americas"), 
   color="red", alpha=0.5, size=4) + theme(legend.position="none") +
  annotate("text", x = c(40000), y = c(73), family="serif", size=6, color="dark blue",
           label = c("United States")) +
  annotate("text", x = c(13000), y = c(63), family="serif", size=6, color="dark green",
           label = c("Venezuela")) +
  annotate("text", x = c(1200), y = c(62), family="serif", size=6, color="dark orange",
           label = c("Haiti")) + theme_tufte()
```

## Boxplots and Dot Plots

* For some plots we have specific geom() e.g. box plots are created with geom_boxplot
* For other plots we can use the geoms we already know, e.g. for dot plots we can use geom_point
* Overall 37 geoms, but good to know a few. Check out the cheatsheet.

Name of Plot  | Geom | Other Features |
--------------|-------|--------------
scatterplot   | point |  |
bubblechart   | point | size mapped to a variable |
barchart      | bar   | |
box-and-whisker plot | boxplot |
line chart    | line | |


## A New Dataset - Organ Donors

```{r}
organs.url <- "https://raw.githubusercontent.com/socviz/soc880/gh-pages/data/organ-procurement.csv"
organs.data <- read.csv(url(organs.url), row.names=1)
```

```{r, get-organdata, echo= TRUE}
organs <- organs.data
dim(organs)
head(organs)
## For convenience, let R know year is a time measure.
organs$year <- as.Date(strptime(organs$year, format="%Y"))
```

```{r}
p <- ggplot(data = organs, aes(x = year, y= donors))
p + geom_point()

```

```{r}
p + geom_line(aes(group = country, color = consent.law)) +
  scale_color_manual(values=c("gray40", "firebrick")) +
  scale_x_date() + 
  labs(x = "Year", y = "Donors", color = "Consent Law") +
  theme(legend.position = "top")
```

### Faceting

```{r}
# ggplot has two faceting functions that do slightly different things: `facet_grid()`, seen here, and `facet_wrap()`. Try them out on the Gapminder data.

p + geom_line(aes(group=country)) +
    labs(x="Year",
         y="Donors") +
    facet_grid(~ consent.law)
```

### A quick bit of data manipulation - Average by group
```{r}
library(dplyr)
by.country <- organs %>% group_by(consent.law, country) %>%
    summarize(donors=mean(donors, na.rm = TRUE))
by.country
```

### Ordered dotplots
```{r}
p <- ggplot(by.country, aes(x=donors, y=country, color=consent.law))
p + geom_point(size=3)
```

### Ordering

```{r}
p <- ggplot(by.country, aes(x=donors, y=reorder(country, donors), 
                            color=consent.law))
p + geom_point(size=3)
```

### Improve the labels
```{r}
p + geom_point(size=3) +
    labs(x="Donor Procurement Rate (per million population)",
         y="", color="Consent Law") +
    theme(legend.position="top")
```


### Alternate Labels...

```{r}
p <- ggplot(by.country, aes(x=donors, y=reorder(country, donors)))
p + geom_point(size=3) +
    facet_grid(consent.law ~ ., scales="free") +
    labs(x="Donor Procurement Rate (per million population)",
         y="",
         color="Consent Law") +
    theme(legend.position="top")
```

### Dotplots
* dotplots are useful to get a summary of one-dimensional categorical data
* Often better than bar or column charts.
* Keeping the categories on the y-axis is often a good idea.
* They also work well to summarize model results with error ranges.

### Boxplots

```{r}
p <- ggplot(data = organs, aes(x = country, y = donors))
p + geom_boxplot() +
  coord_flip() +
  labs(x="", y="Donor Procurement Rate")
```








